\sysfull

### 2017-08-11 paper/background.tex
merge commit
stash pop merge

Energy harvesting devices operate using energy extracted from sources such as RF and solar energy. Energy-harvesting devices do not use a tethered power supply or battery and instead, typically collect energy into a (super-)capacitor, operate when sufficient energy has accumulated, and upon depleting the stored energy, turn off and wait to collect more energy.
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
Many platforms enable intermittent, battery-less energy harvesting-based computation. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other intermittently-powered platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on energy harvesting device executes {\em intermittently} because energy is only available between when the capacitor reaches its threshold and when it is depleted. An intermittent execution is composed of periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of power failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods (but requires longer charge times). Energy-harvesters provide input power that is orders of magnitude less than a device's operating power, making recharging negligible during operation.

Software behaves uniquely when executed intermittently. Each power failure clears volatile memory (registers, stack, and global variables). Non-volatile memory (e.g., FRAM) persists across power failures. On a power failure, control flows back to a prior point in the execution. By default, after a power failure control flows to the beginning of {\tt main()}. Early architectures preserved intermittent progress by periodically capturing and restoring to a non-volatile checkpoint of the volatile execution context, i.e., registers and stack~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.
%
%Even with the checkpoint at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of a WAR dependence between the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an execution of the code in which power fails during the second loop iteration. The problem arises because line 6 of the execution writes $\texttt{Rmnd}$ in the second iteration, but power fails before capturing the next checkpoint. After restarting from the checkpoint, the update to $\texttt{data}$ at line 9 reads the value of $\texttt{Rmnd}$ that was updated before the power failure, producing an incorrect result.
%
As the figure illustrates, checkpoint-based systems risk violating the volatile and non-volatile memory consistency~\cite{dino}. To maintain consistency, versioning systems~\cite{dino,ratchet} {\em version} a subset of non-volatile variables with the checkpointed execution context. Task-based systems~\cite{chain,alpaca}, which we focus on this work and discuss in detail next, ensure state consistency by using programmer, compiler, and runtime support.

@@

Task-based execution models~\cite{dino,chain,alpaca} ask the programmer to decompose their program into a collection tasks. A task is a function with no caller that contains arbitrary computation, sensing, and communication. A programmer describes the sequence in which tasks may execute, creating a task graph. The flow of control from task to task happens at programmer demarcated points and may conditionally depend on program values. Task-based programming abstractions guarantee that tasks exhibit behavior as though they executed {\em atomically}, regardless of arbitrary power failures. Task-based runtime systems ensure task-atomic semantics by ensuring that repeated task executions are idempotent. The key to idempotence is ensuring that updates to non-volatile state made by an interrupted, partial execution of a task are never visible to a subsequent (potentially completing) re-execution of the task.

There are several run time strategies to ensuring task idempotence. One approach, utilized by Alpaca~\cite{alpaca}, is to identify non-volatile data involved in WAR dependences in a task (like DINO~\cite{dino} and Ratchet~\cite{ratchet} did for checkpoints), execute the task using buffered copies of those data, and commit the buffered copies before completing the task. Another strategy is to statically create multiple versions of non-volatile data shared by tasks and ensure by construction that no task reads and writes the same non-volatile memory location~\cite{chain}. Regardless of the idempotence strategy, the key to task-based execution is that fixed, statically-defined tasks execute as though atomic, completing after zero or more attempts.

@@

Despite valuable features, task-based models incur costs that are the motivation for \sys.

\noindent \textbf{Memory Consistency Enforcement Overhead:} Intermittent execution models incur an overhead because they must checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares the overheads for several recent intermittent execution models (generalizing discussion from~\cite[Sec. 2.4]{alpaca}.) The mechanism responsible for the overhead of each model varies, but all the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves volatile memory and registers to and from non-volatile memory, channel accesses directly manipulate non-volatile data~\cite{chain}, privatization copies data from and commits data to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} require all of memory to be non-volatile.

@@

To concretely understand the cost of these manipulations of non-volatile memory, we directly measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations for the MSP430FR5969~\cite{msp430datasheet}. We used EDB~\cite{edb} to monitor the voltage drop on an energy buffering capacitor that powers the device across 1600 read and write accesses to each type of memory, randomizing locations to avoid caching effects. Figure~\ref{fig:framEnergy} summarizes these measurements, illustrating that a non-volatile memory access consumes around 1.5 times as much energy as a volatile memory access in this device. In an intermittently-powered device, the increased energy corresponds to decreased run time between power failures. The overheads imposed by relying on non-volatile memory motivate \sys, which uses memory virtualization to operate primarily using volatile SRAM.

@@

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|}
		\hline
		Platform name & Storage capacitor size \\
		\hline \hline
		%Moo~\cite{moo} & 0.1\,F \\
		WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
		%NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
		NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
		WISP~\cite{wisp5} & 47\,$\mu$F \\
		{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
		{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
		\hline
	\end{tabular} 
	\caption{{\em Default} energy storage sizes for example battery-less platforms. Observe a huge variation in storage capacity.}
	%We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.
	\label{table:capacitor}
\end{table}

\noindent \textbf{Fixed-size Tasks are Inflexible and Inefficient:} If a programmer writes a task that consumes more energy than the device can buffer, the task will never be able to complete using the buffered energy. The consequence of such tasks is a non-progressing program that repeatedly attempts to re-execute the task of excess energy cost. If a programmer writes a program that consumes less energy than the device can buffer, the system may operate inefficiently. When a task and its successor both complete, the two tasks were interrupted by a task boundary that preserves progress and state (i.e., checkpoint or commit). If the two tasks were not also interleaved by a power failure, then the computation done to preserve state was {\em unnecessary}, only incurring an overhead on the tasks' executions. Avoiding excessively costly, non-terminating tasks and short, high-overhead tasks is a programming challenge, given fixed hardware with a fixed energy buffering capacity that might be \emph{different} for each transiently-powered device (see Table~\ref{table:capacitor}). Such tasks also make it extremely difficult to {\em port} code from one device to another. A task that is excessively costly on a device with a small energy buffer may become a short, high-overhead task on a device with a larger energy buffer and vice versa. The key problem is that the programmer writes tasks statically for an energy buffer of a fixed size. \sys's {\em task coalescing} mechanism, which we will describe in Section~\ref{sec:task_coalescing}, is motivated by the above observations.

>>>>

Energy harvesting devices operate using energy extracted from sources such as RF and solar energy. These devices elide tethered power or a battery, instead collecting energy into a capacitor, operating when sufficient energy accumulates, and upon depleting the energy, turning off and recharging.
%
%Batteryless operation has a number of important advantages, making intermittent computing an important research domain. Supplying power to billions~\cite{gartner_iot} of embedded computers using batteries is not sustainable. The European Commission estimates that more than 160 kilotons of consumer batteries enter the European Union annually~\cite{eu_batteries_2016}. Batteries are an environmental risk, are fragile, are limited in their number of charge/discharge cycles, and may require costly physical maintenance that is difficult or impossible deployed (e.g. in space~\cite{kicksat}). By contrast, super-capacitors are durable, promising millions of charge/discharge cycles~\cite[Sec. I]{ongaro_pwre_2012}. The main limitation in moving to capacitor-based energy storage is that capacitor energy density---and consequently operating discharge time---is orders of magnitude less than a battery. 
%
%Given current technology development, battery-less systems are best suited for very long-term sensing and monitoring where access to recharge is either prohibitive or impossible. These include battery-less image capture and processing~\cite{naderiparizi_rfid_2015}, animal monitoring~\cite{thomas_jbcs_2012} or implantable~\cite{rodriguez_tbcs_2015} and digestible~\cite{nadeau_naturebio_2017} sensors.
%
Many platforms enable intermittent, battery-less energy harvesting-based computation. For instance, computational RFIDs---open-source TI MSP430-based~\cite{wolverine} WISP~\cite{wisp5} (with its variants such as WISPCam~\cite{naderiparizi_rfid_2015}, NFC-WISP~\cite{zhao_rfid_2015} or NeuralWISP~\cite{holleman_biocas_2008}), Moo~\cite{moo}, and commercial ones such as~\cite{medusa_farsens_2017}. Other intermittently-powered platforms include ambient backscatter tag~\cite{liu_sigcomm_2013,parks_sigcomm_2014} or battery-less phone~\cite{talla_imwut_2017}.

@@

%Energy-harvesting devices execute software according to the {\em intermittent execution model}. Physically, a device charges until a threshold, operates briefly until its energy is depleted, shuts down recharges, and repeats the cycle.
%
Software running on an energy-harvesting device executes {\em intermittently} because buffered energy is only sometimes available. An intermittent execution is composed of operating periods interspersed with power failures~\cite{dino,chain,alpaca,ratchet}. The frequency of failures depends on the size of the device's energy storage buffer: a larger buffer allows longer operating periods. Energy-harvesters provide input power orders of magnitude less than operating power, making recharging negligible during operation.

Intermittent execution is different from continuous execution. A power failure clears volatile  state (registers, stack, and globals) and non-volatile memory (e.g., FRAM) persists. At a failure, control flows to a prior point in the execution: by default, to the beginning of {\tt main()}. Early intermittent systems preserved progress by periodically checkpointing volatile execution context to non-volatile memory~\cite{mementos,quickrecall}, sometimes using hardware support~\cite{mementos,mottola2017harvos,hibernusplusplus,hibernus,idetic}.

@@

Figure~\ref{fig:code_demo_incosistency} illustrates how state can become inconsistent in an intermittent execution using the cyclic redundancy check (CRC) code from MIBench~\cite{hicks_mibench2_2016}. The code computes the CRC for an $\texttt{nBytes}$ byte message $\texttt{msg}$ with remainder $\texttt{Rmnd}$.

Despite checkpointing at line 3 in Fig.~\ref{fig:crc_example}, the code may compute  $\texttt{data}$ incorrectly because of the read of $\texttt{Rmnd}$ on line 4 and the write of $\texttt{Rmnd}$ on line 5. Figure~\ref{fig:crc_example_war} shows an intermittent execution. The second iteration writes $\texttt{Rmnd}$, but power fails before the checkpoint. After restarting, line 9 reads the updated value of $\texttt{Rmnd}$, producing an incorrect $\texttt{data}$ value.
%
Checkpoint-based systems risk violating memory consistency~\cite{dino}. To maintain consistency, prior systems~\cite{dino,ratchet} {\em version} a subset of non-volatile data with the checkpoint. Task-based systems~\cite{chain,alpaca}, which we focus on, ensure consistency using programmer, compiler, and runtime support.

@@

Task-based execution models~\cite{dino,chain,alpaca} ask the programmer to decompose their program into tasks. A task is a function with no caller containing arbitrary computation, sensing, and communication. A programmer describes task control-flow as a {\em task graph}. Task flow happens at programmer demarcated transition points that may be conditional on program values. Task-based programming abstractions guarantee that tasks execute {\em atomically}, regardless of power failures. Task-based runtime systems ensure task-atomic semantics by ensuring that repeated task executions are idempotent. The key to idempotence is ensuring that non-volatile updates made by an interrupted task are never visible to a future task execution.

There are several run time strategies to ensuring task idempotence. One approach~\cite{alpaca}, is to identify non-volatile data involved in WAR dependences in a task (like DINO~\cite{dino} and Ratchet~\cite{ratchet} did for checkpoints), execute the task using private copies of those data, and commit the private copies on task completion. Another way is to statically create multiple versions of non-volatile data shared by tasks and ensure that no task reads and writes the same version~\cite{chain}. Regardless of the strategy, task-based systems execute statically-defined tasks atomically, completing in one or more attempt.

@@

Task-based models incur costs that motivate \sys.

\noindent \textbf{Memory Consistency Enforcement Overhead:} Intermittent execution models incur overheads to checkpoint data~\cite{dino,ratchet,quickrecall,mementos}, manage channels~\cite{chain}, or privatize and commit data~\cite{alpaca}. Table~\ref{table:chechpoint_comparison} compares overheads for recent intermittent execution models (cf. \cite[Sec. 2.4]{alpaca}.) The mechanism responsible for overhead in each model varies, but the bulk of overhead in all models is a manipulating non-volatile memory. Checkpointing moves data to and from non-volatile memory, channel accesses manipulate non-volatile data~\cite{chain}, privatization copies from and commits to non-volatile memory~\cite{alpaca}, and idempotence solutions~\cite{ratchet} use {\em only} non-volatile memory.

@@

To understand these non-volatile memory overheads, we measured the energy consumption of volatile (SRAM) and non-volatile (FRAM) memory operations for the MSP430FR5969~\cite{msp430datasheet} MCU. We used EDB~\cite{edb} to monitor voltage drop on a capacitor powering the device across 1600 read and write accesses to each memory type, randomizing to avoid caching. Figure~\ref{fig:framEnergy} summarizes the data, showing that a non-volatile memory access consumes around 1.5 times the energy of a volatile access in this device. In an intermittent device, energy cost corresponds to decreased run time between power failures. The cost of non-volatile memory motivates \sys, which virtualizes memory to primarily use SRAM.

@@

%\begin{table}
%	\centering
%	\footnotesize
%	\begin{tabular}{|c|c|}
%		\hline
%		Platform name & Storage capacitor size \\
%		\hline \hline
%		%Moo~\cite{moo} & 0.1\,F \\
%		WISPCam~\cite{naderiparizi_rfid_2015} & 6.08\,mF \\ %tested [11.24, 17.45, 21.98]\,mF
%		%NFC-WISP~\cite{zhao_rfid_2015} & 300\,$\mu$F \\
%		NeuralWISP~\cite{holleman_biocas_2008} & 100\,$\mu$F \\
%		WISP~\cite{wisp5} & 47\,$\mu$F \\
%		{\em BioImpedance} sensor~\cite{rodriguez_tbcs_2015} & 20\,$\mu$F \\
%		{\em Ingestible} sensor~\cite{nadeau_naturebio_2017} & 220\,$\mu$F\\
%		\hline
%	\end{tabular} 
%	\caption{{\em Default} energy storage sizes for example battery-less platforms. Observe a huge variation in storage capacity.}
%	%We note that values for other representative platforms~\cite{medusa_farsens_2017,talla_imwut_2017,liu_sigcomm_2013,parks_sigcomm_2014} were not reported in their respective papers.
%	\label{table:capacitor}
%\end{table}

\noindent \textbf{Fixed-size Tasks are Inflexible and Inefficient:} If a
programmer writes a task that consumes {\em more} energy than the device can
buffer, the task will never complete using buffered energy. {\em preventing
forward progress} by repeatedly re-executing. If a programmer writes tasks that
consume far {\em less} energy than a device can buffer, the system may operate
{\em inefficiently}. When a task and its successor both complete, the two tasks
were interleaved by a task boundary that preserves progress and state (i.e.,
checkpoint or commit). However, absent a power failure, the work to preserve
state was {\em unnecessary}, incurring overhead on the tasks' executions.
Avoiding excessively costly, non-terminating tasks and short, high-overhead
tasks is a programming challenge, given fixed hardware with a fixed energy
buffering capacity. Buffer sizes from prior work vary widely from $20\mu F$~\cite{rodriguez_tbcs_2015}
to $0.1F$~\cite{moo}.
Sizing tasks complicates 
{\em porting} code from one device to another. An excessively costly task 
on a device with a small energy buffer may be a relatively short,
task on a device with a larger buffer. The
key problem is that using existing systems, the programmer statically sizes tasks for a fixed energy buffer. \sys's {\em task coalescing}, described
in Section~\ref{sec:task_coalescing}, is motivated by these observations.
